---
title: "The Paper's Empirical Applications"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Paper's Empirical Applications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5)
has_nsw  <- requireNamespace("qte", quietly = TRUE)
```

```{r lib}
library(PlaceboLM)
```

This vignette reproduces the two empirical applications in Sections 3.2 and 3.3
of Rohde and Hazlett. Both need data that is not bundled with the package;
`data-raw/` contains scripts that fetch and prepare it. Where the data is
absent the code is shown but not run, and the paper's published figures are
quoted so you can check your own run against them.

---

# 1. Job training and earnings (NSW)

The National Supported Work Demonstration randomly assigned disadvantaged
workers to a job-training programme. The experimental benchmark — the number a
good observational method should recover — is a **\$1,794** increase in 1978
earnings, or **\$1,671** once observed covariates are included.

The analysis uses the *non-experimental* comparison sample, in which the
randomised controls are replaced by respondents from the PSID. Regression on
observables famously gives a large negative estimate on this dataset.

The placebo outcome is **1975 earnings** (`re75`) — a pre-treatment measure of
the outcome, which the treatment cannot have affected. So the placebo
imperfection is credibly zero, and the whole argument reduces to reasoning
about relative confounding.

```{r nsw-fit, eval = has_nsw}
data(lalonde, package = "qte")

fit_nsw <- placebo_lm(
  data       = lalonde.psid,
  outcome    = "re78",
  treatment  = "treat",
  placebo    = "re75",
  covariates = c("age", "education", "black", "hispanic",
                 "married", "nodegree"),
  structure  = "placebo_outcome"
)

plm_benchmarks(fit_nsw)
```

The paper reports, for this specification:

| Quantity | Paper |
|---|---|
| Adjusting for observables (`k = 0`) | below −\$5,000 |
| Conventional DID (`m = 1`) | **\$2,087** |
| Equiconfounding after rescaling (`k = 1`) | **\$3,428** |
| Scale factor `SF` | 1.167 |

## Reasoning about `k`

The paper's argument runs as follows. The confounding in the
treatment–1978-earnings relationship should share a sign with the confounding
in the treatment–1975-earnings relationship, so `k > 0`. Both are measures of
the same person's income a few years apart, so we might start near `0.5 < k <
1.5`. But people select into job training on the basis of *recent* income loss
— the Ashenfelter dip — so confounding should be stronger closer to enrolment
in 1975 than in 1978. That gives `0 < k < 1`, with the most weight on
`0.5 < k < 1`.

```{r nsw-bounds, eval = has_nsw}
plm_bounds(fit_nsw, k = c(0.5, 1), n_boot = 500)
```

The paper reports this bound as **[−\$1,249, +\$3,428]**, which contains the
experimental benchmark of \$1,671.

## Backing out the true `k`

Because an experimental benchmark exists here, we can invert the problem and
ask what `k` actually held:

```{r nsw-solve, eval = has_nsw}
plm_solve(fit_nsw, target = 1671)
```

The paper reports **`k = 0.812`** — comfortably inside the `0.5 < k < 1` range
that was argued for *ex ante*, which is the paper's main point about this
application. It also notes that because incomes grew between 1975 and 1978
(`SF = 1.167`), the corresponding raw bias ratio is `m = 0.948`. Conventional
DID assumes `m = 1`, so it lands close to the benchmark here — but by
coincidence of two offsetting violations, not because its assumption was
defensible.

```{r nsw-line, eval = has_nsw, fig.alt = "NSW line plot"}
plm_line_plot(fit_nsw, k = c(0, 2), n_boot = 500)
```

## A deliberately badly scaled placebo

The paper also uses 1975 *unemployment* — a binary variable — as a placebo for
1978 earnings in dollars. The scale factor exceeds 40,000. This is the case
that shows why the package carries both parameterizations: `m` is meaningless
across such different scales, while `k` remains interpretable. The paper finds
the true `k` here is 0.33.

```{r nsw-u, eval = FALSE}
fit_u <- placebo_lm(
  data       = lalonde.psid,
  outcome    = "re78",
  treatment  = "treat",
  placebo    = "u75",
  covariates = c("age", "education", "black", "hispanic",
                 "married", "nodegree"),
  structure  = "placebo_outcome"
)
fit_u$SF                      # very large
plm_bounds(fit_u, k = c(0, 1), n_boot = 500)
```

## Triangulation

Repeating the analysis with 1974 earnings as the placebo gives
**[−\$1,406, +\$3,115]**. Intersecting that with the 1975 result gives
**[−\$1,249, +\$3,115]** — a tighter interval that still rests only on
assumptions the investigator was willing to defend for each placebo
separately.

---

# 2. Zika and birth rates in Brazil

In 2015 the Brazilian state of Pernambuco saw high rates of Zika infection,
while Rio Grande do Sul reported none. Treating the former as treated and the
latter as control, the outcome is the municipality-level 2016 birth rate per
1,000 people and the placebo outcome is the same measure in 2014. As with NSW,
the placebo is pre-treatment, so its imperfection is credibly zero.

```{r zika, eval = FALSE}
# See data-raw/zika.R for acquisition from the Harvard Dataverse
# (Amorim 2022, doi:10.7910/DVN/ENG0IY)
zika <- readRDS("zika.rds")

fit_zika <- placebo_lm(
  data      = zika,
  outcome   = "birth_rate_2016",
  treatment = "treated",
  placebo   = "birth_rate_2014",
  structure = "placebo_outcome"
)

plm_benchmarks(fit_zika)
plm_solve(fit_zika, target = 0)          # where the sign flips
plm_line_plot(fit_zika, k = c(0, 2), n_boot = 500)
```

The paper reports:

| Quantity | Paper |
|---|---|
| Naive difference in means | **+3.4** births per 1,000 |
| DID (`m = 1`) and `k = 1` | **−1.2 to −1.3** |
| Sign flips at | **`k ≈ 0.7`** |
| Estimate at `k = 1.2` | **−2.0** |

Here, unlike NSW, there is little reason to argue `k < 1`: selection into
treatment was close in time to the post-treatment outcome. So `k > 1` is if
anything the more plausible region, and over most of the *ex ante* plausible
range the data are consistent with Zika having **reduced** births — the
opposite sign to the naive comparison. Anyone arguing Zika did not reduce
births has to defend `k < 0.7`.

---

## Reference

Rohde, A. and Hazlett, C. Causal progress with imperfect placebo treatments and
outcomes.

Amorim, L. (2022). Replication Data for: Zika Epidemic and Birth Rates in
Brazil. <https://doi.org/10.7910/DVN/ENG0IY>

LaLonde, R. J. (1986). Evaluating the econometric evaluations of training
programmes with experimental data. *American Economic Review*, 76(4), 604–620.
