---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5)
```

This vignette works through the placebo-treatment simulation of Section 3.1 of
Rohde and Hazlett, which is small enough to run here and has a known right
answer to check against.

## The problem

We want the effect of regular exercise on heart disease. We worry that
socioeconomic status confounds it: it plausibly affects both the ability to
exercise and, through diet, healthcare access, and stress, the risk of heart
disease. In a real study we would not observe it.

We do observe whether people take supplemental vitamins. Vitamin use is
plausibly driven by the same health-seeking behaviour and resources that drive
exercise, but we expect it to have little or no effect on heart disease. That
makes it a **placebo treatment**: a variable that shares confounders with the
real treatment but does not itself cause the outcome.

The idea of the method is to use the confounding we *can* see — in the
vitamins–heart-disease relationship — to reason about the confounding we
cannot, in the exercise–heart-disease relationship.

## Simulating the data

This is the data-generating process from the paper's appendix, unchanged.

```{r dgp}
library(PlaceboLM)

set.seed(1)
n <- 5000

ses  <- runif(n, 0, 1)                                  # socioeconomic status
phy  <- rbinom(n, size = 1, prob = pnorm(2 * ses))      # regular exercise
diet <- rbinom(n, size = 1, prob = pnorm(-1 + ses))     # vitamin supplements
HD   <- rbinom(n, size = 1, prob = pnorm(-2 + -0.4 * phy + 4 * ses))

health_data <- data.frame(
  SES                 = ses,
  Physical_Activity   = phy,
  Dietary_Supplements = diet,
  Heart_Disease       = HD
)

round(c(exercise = mean(phy), vitamins = mean(diet),
        heart_disease = mean(HD)), 3)
```

Because this is a simulation we can compute the answer an investigator could
not: the coefficient on exercise in the *long* regression that includes `SES`.

```{r truth}
truth <- coef(lm(Heart_Disease ~ Physical_Activity + Dietary_Supplements + SES,
                 data = health_data))[["Physical_Activity"]]
round(truth, 4)
```

Roughly a 10-percentage-point reduction in the probability of heart disease.
Everything below is an attempt to get near that number without using `SES`.

## Fitting

```{r fit}
fit <- placebo_lm(
  data      = health_data,
  outcome   = "Heart_Disease",
  treatment = "Physical_Activity",
  placebo   = "Dietary_Supplements",
  structure = "placebo_treatment"
)

fit
```

The printout states the **estimand** — the coefficient on exercise in the long
regression that would include socioeconomic status if we could observe it. That
is a linear projection. It is an average causal effect only under further
assumptions, such as no effect heterogeneity, so the package calls it
`adjusted_coefficient` rather than `estimate`. It also lists the assumptions
this structure commits you to, and the two sensitivity parameters the analysis
rests on.

`beta_Y_P_given_DZX` is the **placebo imperfection**: the effect vitamins would
still have on heart disease after adjusting for exercise and for the unobserved
confounders. We are willing to argue this is zero — that is the substantive
claim that vitamins do not prevent heart disease.

The other is relative confounding, and it comes in two forms.

## Two ways to talk about relative confounding

`m` is the raw ratio of the bias in the exercise–heart-disease relationship to
the bias in the vitamins–heart-disease relationship. Equiconfounding on the raw
scale is `m = 1`.

The paper computes the `m = 1` estimate here and calls it the conventional DID
estimate, by analogy with how difference-in-differences arithmetic works. Worth
being precise about what that does and does not mean: the *interpretation* of
`m = 1` as parallel trends belongs to the case where the placebo is a
pre-treatment measure of the outcome. This example uses a placebo treatment, so
`m = 1` is the same arithmetic without that reading, and the benchmark label
reflects it.

`k` is the same idea after correcting for the fact that exercise and vitamin
use are on different scales, related by `m = k * SF` where `SF` is the scale factor
reported above. Reasoning about `k` asks: *is exercise more or less confounded
with heart disease than vitamin use is?*

Supply whichever you prefer:

```{r params}
plm_estimate(fit, k = 0)   # no unobserved confounding at all
plm_estimate(fit, m = 1)   # equiconfounding on the raw scale
plm_estimate(fit, k = 1)   # equal confounding, after rescaling
```

These three are worth having together, so they have their own function:

```{r benchmarks}
plm_benchmarks(fit)
```

Assuming away unobserved confounding gives roughly **+17 percentage points** —
exercise appears to *cause* heart disease, which is not credible and is a good
illustration of what confounding does here. The two placebo-adjusted benchmarks
give about **+1pp** (raw-scale equiconfounding, `m = 1`, which here corresponds
to `k = 0.85`) and **−1pp** (equiconfounding after rescaling, `k = 1`). Both are far closer to the truth of
about −10pp than assuming no unobserved confounding, though neither is a good
approximation to it — which is the paper's point: these are still *point*
assumptions.

## Reasoning about a range

The point of the method is that we do not have to commit to a single value.
The paper's reasoning for this example runs roughly: the biases plausibly share
a sign, so `k > 0`; exercise probably depends on socioeconomic status *more*
strongly than cheap vitamin supplements do, so probably `k > 1`; but it is hard
to argue for `k > 3`.

That range of assumptions maps to a range of estimates:

```{r bounds}
plm_bounds(fit, k = c(0, 3), n_boot = 200)
```

The truth is inside it. Note what this bought us: we did not have to defend a
point assumption, only an interval, and the interval was wide enough to be
defensible.

`lower` and `upper` are the bounds themselves. The `*_boot_q` columns summarise
sampling variability in those bounds; they are not a confidence interval for the
identified set, and `?plm_bounds` spells out the distinction.

```{r line, fig.alt = "Adjusted estimate against k"}
plm_line_plot(fit, k = c(0, 3), n_boot = 200)
```

## Where does the conclusion flip?

The inverse question is often the more useful one. Rather than asking what
estimate a given `k` implies, ask which `k` would overturn a conclusion:

```{r solve}
plm_solve(fit, target = 0)
```

Anyone who wants to claim exercise does *not* reduce heart disease has to argue
for a `k` below that value — that exercise is *less* confounded with heart
disease than vitamin use is. In this setting that is a hard case to make, which
is exactly the kind of statement the method is designed to produce.

Since this is a simulation, we can also check what the true `k` was:

```{r truek}
plm_solve(fit, target = truth, imperfection = 0)
```

The paper reports this as **`k = 1.45`**: the confounding between vitamin use
and heart disease would have to be about 1.45 times as severe as that between
exercise and heart disease. An investigator willing to entertain that value in
their range would have covered the truth — and the range argued for above,
`0 ≤ k ≤ 3`, does.

## Two parameters at once

If the placebo is not perfect, both parameters vary, and the contour plot shows
the whole surface. The red contour is the set of assumptions at which the sign
of the estimate flips.

```{r contour, fig.alt = "Contour of the adjusted estimate"}
plm_contour_plot(fit, k = c(0, 3))
```

## Where to go next

* `vignette("structures")` — the five causal structures the package supports,
  and how to choose between them.
* `vignette("applications")` — the paper's two empirical applications.

## Reference

Rohde, A. and Hazlett, C. Causal progress with imperfect placebo treatments and
outcomes.
