---
title: "Getting Started: The LaLonde Job-Training Example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started: The LaLonde Job-Training Example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)
set.seed(0)
```

## Background

This vignette replicates the LaLonde (1986) job-training analysis from
Rohde and Hazlett (2025). The goal is to estimate the effect of a job-training
programme (`treat`) on 1978 earnings (`re78`), using 1974 pre-treatment
earnings (`re74`) as a *placebo outcome*. The data come from the PSID
comparison sample.

The core insight of the PlaceboLM framework is that pre-treatment outcomes
are imperfect placebos. They share unobserved confounders with the
treatment-outcome relationship, but the confounding may not be *equal* across
the two periods. The sensitivity parameters `k` and `coef_P_D_given_XZ`
let us reason about how far the equiconfounding assumption can be relaxed
before the qualitative conclusion changes.

## Data

```{r data}
library(PlaceboLM)

data(lalonde, package = "qte")
lalonde_psid <- lalonde.psid

head(lalonde_psid[, c("re78", "treat", "re74",
                       "age", "education", "black",
                       "hispanic", "married", "nodegree")])
```

## Setting up the placebo model

We specify:

- **Outcome** `re78`: 1978 earnings (post-treatment).
- **Treatment** `treat`: indicator for assignment to the job-training programme.
- **Placebo outcome** `re74`: 1974 earnings (pre-treatment, before the programme).
- **Causal structure** `DP = ""`, `PY = ""`: we assume the pre-treatment
  outcome has no direct relationship with either the treatment or the
  post-treatment outcome beyond shared confounders.  This is the standard
  parallel-trends-style assumption.
- **Partial-identification parameters**: `k` ∈ [−2, 2] scales the
  confounding in `re78` relative to the confounding in `re74`; `coef_P_D_given_XZ`
  is the residual association between `re74` and `treat` after removing the
  influence of all observed *and* unobserved confounders.

```{r setup-plm}
plm <- placeboLM(
  data                  = lalonde_psid,
  outcome               = "re78",
  treatment             = "treat",
  placebo_outcome       = "re74",
  observed_covariates   = c("age", "education", "black",
                             "hispanic", "married", "nodegree"),
  partialIDparam_minmax = list(
    k                   = c(-2, 2),
    coef_P_D_given_XZ   = c(-15000, 15000)
  )
)
```

The printed messages confirm the placebo type and the two regression
specifications that will be used.

## Summary table

`placeboLM_table()` always reports three benchmark estimates first:

| Row | Meaning |
|---|---|
| **No Unobserved Confounding** | OLS estimate; assumes away all unobserved confounding (`k = 0`). |
| **DID** | The value of `k` at which the estimate equals a difference-in-differences benchmark; equiconfounding on the observed scale. |
| **Perfect Placebo (k=1)** | Assumes unobserved confounding is *identical* (on the regression scale) across the two periods. |

Below these, a grid of estimates at the 25th, 50th, and 75th percentiles of
each parameter range is printed.

```{r table, cache = TRUE}
placeboLM_table(
  plm,
  n_boot   = 200,
  ptiles   = c(0.25, 0.5, 0.75),
  alpha    = 0.05,
  decimals = 0
)
```

**Reading the table.** The SOO estimate is around −5 900, suggesting a
negative effect if we assume no unobserved confounding — implausible given
the non-random assignment. The DID benchmark (equiconfounding) produces an
estimate near 1 700, consistent with the experimental benchmark of roughly
1 800. The perfect-placebo estimate (`k = 1`) is around 3 100.

## Contour plot

The contour plot shows the estimate as a continuous function of both
sensitivity parameters simultaneously. The zero contour (red) is the
**tipping line**: parameter combinations above (or below) it are the ones
that would change the sign of the estimated effect.

```{r contour, cache = TRUE}
placeboLM_contour_plot(plm, gran = 60, decimals = 0)
```

The green square marks the DID benchmark, the blue triangle the perfect-placebo
benchmark (`k = 1`, `coef_P_D_given_XZ = 0`), and the navy diamond the SOO
estimate. The DID benchmark lies well above the zero contour, indicating that
equiconfounding is sufficient to recover a positive effect.

## Line plots

Line plots slice through the two-dimensional parameter space, fixing
`coef_P_D_given_XZ` at three levels (0th, 50th, and 100th percentiles of its
range) and varying `k` along the x-axis. Shaded bands are 95 % bootstrap
confidence intervals.

```{r line, cache = TRUE}
placeboLM_line_plot(
  plm,
  bootstrap   = TRUE,
  n_boot      = 200,
  ptiles      = c(0, 0.5, 1),
  focus_param = "k",
  ptile_param = "coef_P_D_given_XZ",
  gran        = 15,
  alpha       = 0.05,
  decimals    = 0
)
```

When `coef_P_D_given_XZ = 0` (the middle panel, and the panel where benchmark
points are drawn), the estimate is positive and statistically distinguishable
from zero for values of `k` near the DID benchmark. The zero-crossing of the
estimate line — the robustness value for `k` — gives the minimum departure from
equiconfounding required to overturn the positive conclusion.

## References

LaLonde, R. J. (1986). Evaluating the econometric evaluations of training
programmes with experimental data. *American Economic Review*, 76(4), 604–620.

Rohde, A. and Hazlett, C. (2025). Causal progress with imperfect placebo
treatments and outcomes. *Journal of the Royal Statistical Society: Series A*.
<https://doi.org/10.1093/jrsssa/qnaf001>
