---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5)
```

This vignette works through the placebo-treatment simulation of Section 3.1 of
Rohde and Hazlett, which is small enough to run here and has a known right
answer to check against.

## The problem

We want the effect of regular exercise (`D`) on heart disease (`Y`). We worry
that socioeconomic status (`Z`) confounds it: it plausibly affects both the
ability to exercise and, through diet, healthcare access, and stress, the risk
of heart disease. We do not observe `Z`.

We do observe whether people take supplemental vitamins (`P`). Vitamin use is
plausibly driven by the same health-seeking behaviour and resources that drive
exercise, but we expect it to have little or no effect on heart disease. That
makes it a **placebo treatment**: a variable that shares confounders with the
real treatment but does not itself cause the outcome.

The idea of the method is to use the confounding we *can* see — in the
vitamins–heart-disease relationship — to reason about the confounding we
cannot, in the exercise–heart-disease relationship.

## Simulating the data

```{r dgp}
library(PlaceboLM)

set.seed(2023)
n <- 5000

Z <- runif(n)                                   # latent socioeconomic status
D <- rbinom(n, 1, pnorm(2 * Z))                 # regular exercise
P <- rbinom(n, 1, pnorm(-1 + Z))                # vitamin supplements
Y <- rbinom(n, 1, pnorm(-2 - 0.43 * D + 4 * Z)) # heart disease

sim <- data.frame(Y = Y, D = D, P = P, Z = Z)
round(c(exercise = mean(D), vitamins = mean(P), heart_disease = mean(Y)), 3)
```

Because this is a simulation we can compute the answer an investigator could
not: the coefficient on `D` in the *long* regression that includes `Z`.

```{r truth}
truth <- coef(lm(Y ~ D + P + Z, data = sim))[["D"]]
round(truth, 4)
```

Roughly a 10-percentage-point reduction in the probability of heart disease.
Everything below is an attempt to get near that number without using `Z`.

## Fitting

```{r fit}
fit <- placebo_lm(
  data      = sim,
  outcome   = "Y",
  treatment = "D",
  placebo   = "P",
  structure = "placebo_treatment"
)

fit
```

The printout names the two sensitivity parameters the analysis rests on.
`beta_Y_P_given_DZX` is the **placebo imperfection**: the effect vitamins would
still have on heart disease after adjusting for exercise and for the unobserved
confounders. We are willing to argue this is zero — that is the substantive
claim that vitamins do not prevent heart disease.

The other is relative confounding, and it comes in two forms.

## Two ways to talk about relative confounding

`m` is the raw ratio of the bias in the exercise–heart-disease relationship to
the bias in the vitamins–heart-disease relationship. Conventional
difference-in-differences is exactly `m = 1`.

`k` is the same idea after correcting for the fact that `D` and `P` are on
different scales, related by `m = k * SF` where `SF` is the scale factor
reported above. Reasoning about `k` asks: *is exercise more or less confounded
with heart disease than vitamin use is?*

Supply whichever you prefer:

```{r params}
plm_estimate(fit, k = 0)   # no unobserved confounding at all
plm_estimate(fit, m = 1)   # conventional difference-in-differences
plm_estimate(fit, k = 1)   # equal confounding, after rescaling
```

These three are worth having together, so they have their own function:

```{r benchmarks}
plm_benchmarks(fit)
```

Assuming away unobserved confounding gives roughly **+17 percentage points** —
exercise appears to *cause* heart disease, which is not credible and is a good
illustration of what confounding does here. Both placebo-adjusted benchmarks
land far closer to the truth of −10pp, though neither hits it.

## Reasoning about a range

The point of the method is that we do not have to commit to a single value.
The paper's reasoning for this example runs roughly: the biases plausibly share
a sign, so `k > 0`; exercise probably depends on socioeconomic status *more*
strongly than cheap vitamin supplements do, so probably `k > 1`; but it is hard
to argue for `k > 3`.

That range of assumptions maps to a range of estimates:

```{r bounds}
plm_bounds(fit, k = c(0, 3), n_boot = 200)
```

The truth is inside it. Note what this bought us: we did not have to defend a
point assumption, only an interval, and the interval was wide enough to be
defensible.

```{r line, fig.alt = "Adjusted estimate against k"}
plm_line_plot(fit, k = c(0, 3), n_boot = 200)
```

## Where does the conclusion flip?

The inverse question is often the more useful one. Rather than asking what
estimate a given `k` implies, ask which `k` would overturn a conclusion:

```{r solve}
plm_solve(fit, target = 0)
```

Anyone who wants to claim exercise does *not* reduce heart disease has to argue
for a `k` below that value — that exercise is *less* confounded with heart
disease than vitamin use is. In this setting that is a hard case to make, which
is exactly the kind of statement the method is designed to produce.

Since this is a simulation, we can also check what the true `k` was:

```{r truek}
oracle <- coef(lm(Y ~ D + P + Z, data = sim))[["P"]]
plm_solve(fit, target = truth, imperfection = oracle)
```

An investigator who was willing to entertain that value in their range would
have covered the truth.

## Two parameters at once

If the placebo is not perfect, both parameters vary, and the contour plot shows
the whole surface. The red contour is the set of assumptions at which the sign
of the estimate flips.

```{r contour, fig.alt = "Contour of the adjusted estimate"}
plm_contour_plot(fit, k = c(0, 3))
```

## Where to go next

* `vignette("structures")` — the five causal structures the package supports,
  and how to choose between them.
* `vignette("applications")` — the paper's two empirical applications.

## Reference

Rohde, A. and Hazlett, C. Causal progress with imperfect placebo treatments and
outcomes.
