---
title: "Placebo Configurations: All Six Single-Placebo Types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Placebo Configurations: All Six Single-Placebo Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Overview

PlaceboLM supports six distinct single-placebo causal structures, selected via
the `DP` and `PY` arguments. This vignette walks through each one using a
common synthetic data-generating process so that the "oracle" answer (the
value of the sensitivity parameter that recovers the true causal effect) is
known.

## Data-generating process

```{r dgp}
library(PlaceboLM)

set.seed(42)
n <- 2000

U <- rnorm(n)          # unobserved confounder
X <- rnorm(n)          # observed covariate

D  <- X + U + rnorm(n)          # treatment
P  <- X + U + rnorm(n)          # placebo (shares confounders with D)
Y  <- D + X + U + rnorm(n)      # outcome  (true effect of D is 1)
N  <- D + X + U + rnorm(n)      # placebo outcome (same structure as Y)

dat <- data.frame(Y = Y, D = D, P = P, N = N, X = X, U = U)
```

The true causal effect of `D` on `Y` is **1**. Both `D` and `P` share the
same unobserved confounder `U`, and the same is true of `N` (placebo outcome).

For each configuration, the "oracle" regression conditions on both `X` and `U`
to reveal the residual placebo association after removing all confounding. In
practice `U` is unobserved; specifying the range of the corresponding
sensitivity parameter is the analyst's responsibility.

---

## 1. No Direct Relationships — Placebo Outcome
**`DP = ""`, `PY = ""`**

The placebo outcome `N` is affected by the same confounders as `Y`, but
neither `D` nor `Y` directly causes `N` (beyond those shared confounders).
This is the DiD-like setting: pre-treatment outcomes used as a placebo.

Estimating equations:
$$\hat\beta_{YD \cdot X} \quad\text{and}\quad \hat\beta_{ND \cdot X}$$

Sensitivity parameter: `coef_P_D_given_XZ` = the coefficient of `D` in
`N ~ D + X + U` (zero means the confounding is fully eliminated by the
placebo relationship).

```{r type1}
plm1 <- placeboLM(
  data                  = dat,
  outcome               = "Y",
  treatment             = "D",
  placebo_outcome       = "N",
  observed_covariates   = "X",
  partialIDparam_minmax = list(k = c(-2, 2), coef_P_D_given_XZ = c(-2, 2))
)

# Oracle: what is coef_P_D_given_XZ when U is observed?
oracle_coef <- coef(lm(N ~ D + X + U, data = dat))[["D"]]
cat("Oracle coef_P_D_given_XZ:", round(oracle_coef, 3), "\n")
cat("PLM estimate at (k=1, oracle coef):",
    round(placeboLM_point_estimate(
      plm1, list(k = 1, coef_P_D_given_XZ = oracle_coef),
      bootstrap = FALSE)[1], 3), "\n")
```

The oracle value is close to zero because `U` explains most of the `D`–`N`
association not accounted for by `X`. At `k = 1` and the oracle coefficient,
the PLM estimate recovers the true effect of 1.

---

## 2. No Direct Relationships — Placebo Treatment
**`DP = ""`, `PY = ""`**

The placebo *treatment* `P` is used instead of a placebo outcome. Fitting
`Y ~ D + P + X` and adjusting for residual confounding from `P`.

```{r type2}
plm2 <- placeboLM(
  data                  = dat,
  outcome               = "Y",
  treatment             = "D",
  placebo_treatment     = "P",
  observed_covariates   = "X",
  partialIDparam_minmax = list(k = c(-2, 2), coef_Y_P_given_DXZ = c(-2, 2))
)

oracle_coef2 <- coef(lm(Y ~ D + P + X + U, data = dat))[["P"]]
cat("Oracle coef_Y_P_given_DXZ:", round(oracle_coef2, 3), "\n")
cat("PLM estimate at (k=1, oracle coef):",
    round(placeboLM_point_estimate(
      plm2, list(k = 1, coef_Y_P_given_DXZ = oracle_coef2),
      bootstrap = FALSE)[1], 3), "\n")
```

---

## 3. Treatment Causes Placebo (D→P)
**`DP = "->"`, `PY = ""`**

`D` has a direct causal effect on the placebo outcome `N`, in addition to the
shared confounders. The adjustment formula still uses the D-coefficient from
`Y ~ D + X` and `N ~ D + X`, but the interpretation of `coef_P_D_given_XZ`
now includes the direct D→N effect.

```{r type3}
# Augment DGP so D causes N directly
N_dp <- N + 0.5 * D    # D→N direct effect of 0.5
dat$N_dp <- N_dp

plm3 <- placeboLM(
  data                  = dat,
  outcome               = "Y",
  treatment             = "D",
  placebo_outcome       = "N_dp",
  DP                    = "->",
  observed_covariates   = "X",
  partialIDparam_minmax = list(k = c(-2, 2), coef_P_D_given_XZ = c(-2, 2))
)

# Oracle residual coefficient: D in N_dp ~ D + X + U
oracle_coef3 <- coef(lm(N_dp ~ D + X + U, data = dat))[["D"]]
cat("Oracle coef_P_D_given_XZ:", round(oracle_coef3, 3), "\n")
cat("PLM estimate at (k=1, oracle coef):",
    round(placeboLM_point_estimate(
      plm3, list(k = 1, coef_P_D_given_XZ = oracle_coef3),
      bootstrap = FALSE)[1], 3), "\n")
```

Notice the oracle coefficient is now larger (≈ 0.5 + confounding component)
relative to the no-direct-path case.

---

## 4. Placebo Causes Outcome (P→Y)

### 4a. Placebo Treatment version
**`DP = ""`, `PY = "->"`**

The placebo treatment `P` directly causes `Y`. The regression `Y ~ D + P + X`
includes `P` to absorb its direct effect; residual confounding in the `P`–`Y`
coefficient is the sensitivity parameter.

```{r type4a}
Y_py <- Y + 0.3 * P    # P→Y direct effect of 0.3
dat$Y_py <- Y_py

plm4a <- placeboLM(
  data                  = dat,
  outcome               = "Y_py",
  treatment             = "D",
  placebo_treatment     = "P",
  PY                    = "->",
  observed_covariates   = "X",
  partialIDparam_minmax = list(k = c(-2, 2), coef_Y_P_given_DXZ = c(-2, 2))
)

oracle_coef4a <- coef(lm(Y_py ~ D + P + X + U, data = dat))[["P"]]
cat("Oracle coef_Y_P_given_DXZ:", round(oracle_coef4a, 3), "\n")
cat("PLM estimate at (k=1, oracle coef):",
    round(placeboLM_point_estimate(
      plm4a, list(k = 1, coef_Y_P_given_DXZ = oracle_coef4a),
      bootstrap = FALSE)[1], 3), "\n")
```

### 4b. Placebo Outcome version
**`DP = ""`, `PY = "->"`**

The placebo outcome `N` directly causes `Y`. The estimating equations now use
`Y ~ D + N + X` and `N ~ D + X`.

```{r type4b}
Y_py2 <- Y + 0.3 * N
dat$Y_py2 <- Y_py2

plm4b <- placeboLM(
  data                  = dat,
  outcome               = "Y_py2",
  treatment             = "D",
  placebo_outcome       = "N",
  PY                    = "->",
  observed_covariates   = "X",
  partialIDparam_minmax = list(k = c(-2, 2), coef_P_D_given_XZ = c(-2, 2))
)

oracle_coef4b <- coef(lm(N ~ D + X + U, data = dat))[["D"]]
cat("Oracle coef_P_D_given_XZ:", round(oracle_coef4b, 3), "\n")
cat("PLM estimate at (k=1, oracle coef):",
    round(placeboLM_point_estimate(
      plm4b, list(k = 1, coef_P_D_given_XZ = oracle_coef4b),
      bootstrap = FALSE)[1], 3), "\n")
```

---

## 5. Placebo is a Mediator (D→P→Y)

### 5a. Placebo Outcome version
**`DP = "->"`, `PY = "->"`**

Both `D→N` and `N→Y` direct effects exist. PlaceboLM targets the **total**
effect of `D` on `Y`. The estimating equations are `Y ~ D + X` and
`N ~ D + X`.

```{r type5a}
plm5a <- placeboLM(
  data                  = dat,
  outcome               = "Y",
  treatment             = "D",
  placebo_outcome       = "N",
  DP                    = "->",
  PY                    = "->",
  observed_covariates   = "X",
  partialIDparam_minmax = list(k = c(-2, 2), coef_P_D_given_XZ = c(-2, 2))
)
```

### 5b. Placebo Treatment version
**`DP = "->"`, `PY = "->"`**

Both `D→P` and `P→Y` hold. Estimating equations: `Y ~ D + X` and
`Y ~ D + P + X`.

```{r type5b}
plm5b <- placeboLM(
  data                  = dat,
  outcome               = "Y",
  treatment             = "D",
  placebo_treatment     = "P",
  DP                    = "->",
  PY                    = "->",
  observed_covariates   = "X",
  partialIDparam_minmax = list(k = c(-2, 2), coef_Y_P_given_DXZ = c(-2, 2))
)
```

---

## 6. Placebo is an Observed Confounder (P→D)
**`DP = "<-"`**

`P` directly causes `D` (it is an observed confounder). Estimating equations:
`Y ~ D + P + X` and `D ~ P + X`. The sensitivity parameter `coef_D_P_given_XZ`
is the residual coefficient of `P` in `D ~ P + X + U`.

```{r type6}
D_conf <- D + 0.4 * P    # P→D direct effect
dat$D_conf <- D_conf

plm6 <- placeboLM(
  data                  = dat,
  outcome               = "Y",
  treatment             = "D_conf",
  placebo_treatment     = "P",
  DP                    = "<-",
  observed_covariates   = "X",
  partialIDparam_minmax = list(k = c(-2, 2), coef_D_P_given_XZ = c(-2, 2))
)

oracle_coef6 <- coef(lm(D_conf ~ P + X + U, data = dat))[["P"]]
cat("Oracle coef_D_P_given_XZ:", round(oracle_coef6, 3), "\n")
cat("PLM estimate at (k=1, oracle coef):",
    round(placeboLM_point_estimate(
      plm6, list(k = 1, coef_D_P_given_XZ = oracle_coef6),
      bootstrap = FALSE)[1], 3), "\n")
```

---

## 7. Outcome Causes Placebo (Y→P)
**`PY = "<-"`**

`Y` has a direct causal effect on the placebo outcome `N` (reverse causation).
Estimating equations: `Y ~ D + X` and `N ~ Y + D + X`. Sensitivity parameter:
`coef_P_Y_given_DXZ`.

```{r type7}
N_yp <- N + 0.4 * Y    # Y→N direct effect
dat$N_yp <- N_yp

plm7 <- placeboLM(
  data                  = dat,
  outcome               = "Y",
  treatment             = "D",
  placebo_outcome       = "N_yp",
  PY                    = "<-",
  observed_covariates   = "X",
  partialIDparam_minmax = list(k = c(-2, 2), coef_P_Y_given_DXZ = c(-2, 2))
)

oracle_coef7 <- coef(lm(N_yp ~ Y + D + X + U, data = dat))[["Y"]]
cat("Oracle coef_P_Y_given_DXZ:", round(oracle_coef7, 3), "\n")
cat("PLM estimate at (k=1, oracle coef):",
    round(placeboLM_point_estimate(
      plm7, list(k = 1, coef_P_Y_given_DXZ = oracle_coef7),
      bootstrap = FALSE)[1], 3), "\n")
```

---

## Summary

| Type | `DP` | `PY` | Placebo kind | Key sensitivity parameter |
|---|---|---|---|---|
| No Direct Relationships | `""` | `""` | outcome or treatment | `k`, `coef_P_D_given_XZ` / `coef_Y_P_given_DXZ` |
| Treatment Causes Placebo | `"->"` | `""` | outcome | `k`, `coef_P_D_given_XZ` |
| Placebo Causes Outcome | `""` | `"->"` | outcome or treatment | `k`, `coef_P_D_given_XZ` / `coef_Y_P_given_DXZ` |
| Mediator | `"->"` | `"->"` | outcome or treatment | `k`, `coef_P_D_given_XZ` / `coef_Y_P_given_DXZ` |
| Observed Confounder | `"<-"` | any | treatment | `k`, `coef_D_P_given_XZ` |
| Outcome Causes Placebo | any | `"<-"` | outcome | `k`, `coef_P_Y_given_DXZ` |

In every case, setting all sensitivity parameters to zero gives the
selection-on-observables (SOO) estimate; setting `k` to its DID value recovers
the difference-in-differences benchmark; and setting `k = 1` with the
second parameter at zero gives the "perfect placebo" estimate.

## References

Rohde, A. and Hazlett, C. (2025). Causal progress with imperfect placebo
treatments and outcomes. *Journal of the Royal Statistical Society: Series A*.
<https://doi.org/10.1093/jrsssa/qnaf001>
