---
title: "Choosing a Causal Structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a Causal Structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 5)
```

```{r lib}
library(PlaceboLM)
```

The algebra behind this package is a property of linear regression and applies
to any pair of regressions. What it *means* depends on the causal structure you
are willing to assume. That structure determines which regressions to run,
which coefficient is the sensitivity parameter, and how to interpret relative
confounding.

## A common data-generating process

Throughout, `U` is an unobserved confounder shared by the treatment, the
placebo, and the outcome; `X` is an observed covariate. The true effect of `D`
on `Y` is 1. Because `U` is in the data frame we can compute each structure's
*oracle* sensitivity value — the number an investigator would have to postulate
to recover the truth.

```{r dgp}
set.seed(20240101)
n <- 4000

U <- rnorm(n); X <- rnorm(n)
D <- X + U + rnorm(n)
P <- X + U + rnorm(n)
Y <- D + X + U + rnorm(n)

dat <- data.frame(Y = Y, D = D, P = P, X = X, U = U)
```

## Two ways to say which structure you are in

You can name the structure directly. It selects among the graphs in Tables 2
and 3 of Rohde and Hazlett:

```{r table}
plm_structure_table()[, c("structure", "paper_ref", "sens_param")]
```

But names like `observed_confounder_1` and `observed_confounder_2` mean nothing
without the paper open, and picking the wrong one produces a plausible number
rather than an error. Usually it is better to state the causal assumptions and
let the package work out which row of the taxonomy that is:

```{r edges-demo}
fit <- placebo_lm(dat, "Y", "D", "P", covariates = "X", edges = "P->D")
```

Edges are written between the *roles* — `D` (treatment), `P` (placebo), `Y`
(outcome) — not your column names. The full mapping:

```{r edge-table}
plm_edge_table()
```

Two edge sets appear twice, and that ambiguity is genuine rather than a gap in
the encoding. With no direct edges, `P` can be read as a placebo outcome or a
placebo treatment. With `P->Y`, the paper explicitly gives two readings of the
same graph (its Table 2[c]): an imperfect placebo treatment, or an observed
confounder handled on the placebo-outcome side. In both cases `placebo_role`
picks the reading, and omitting it is an error rather than a silent default:

```{r ambiguous, error = TRUE}
placebo_lm(dat, "Y", "D", "P", covariates = "X", edges = "P->Y")
```

## Placebo outcome

`P` shares confounders with `Y` but is not caused by `D`. The canonical case is
a pre-treatment measure of the outcome, which is how difference-in-differences
fits into this framework.

```{r po}
po <- placebo_lm(dat, "Y", "D", "P", covariates = "X",
                 structure = "placebo_outcome")
po$formulas
```

The sensitivity parameter is the residual `P`–`D` association: how much the
treatment still moves the placebo once the unobserved confounders are
accounted for. Zero means a perfect placebo. Here is the oracle value and the
estimate it produces:

```{r po-oracle}
oracle_po <- coef(lm(P ~ D + X + U, data = dat))[["D"]]
c(oracle       = oracle_po,
  adjusted     = plm_estimate(po, k = 1, imperfection = oracle_po),
  unadjusted   = plm_estimate(po, k = 0))
```

The adjusted estimate recovers the true effect of 1; assuming no unobserved
confounding does not.

## Placebo treatment

`P` shares confounders with `D` but does not cause `Y`. Both go into a single
regression, and the sensitivity parameter is the residual `P`–`Y` association.

This structure additionally requires that `P` is not a descendant of `D` and
`D` is not a descendant of `P`, and that each is a good or neutral control for
the other. If you are unsure, draw the graph.

```{r pt}
pt <- placebo_lm(dat, "Y", "D", "P", covariates = "X",
                 structure = "placebo_treatment")
pt$formulas

oracle_pt <- coef(lm(Y ~ D + P + X + U, data = dat))[["P"]]
c(oracle   = oracle_pt,
  adjusted = plm_estimate(pt, k = 1, imperfection = oracle_pt))
```

## Observed confounder 1

`P` causes `Y` and shares confounders with `D`. You could treat `P` as a
placebo outcome, but then the confounding path `D <- Z -> P -> Y` appears in
both the numerator and denominator of `k`, which makes it hard to interpret.
The fix is to put `P` into the outcome regression, where it blocks that path.

```{r oc1}
oc1 <- placebo_lm(dat, "Y", "D", "P", covariates = "X",
                  structure = "observed_confounder_1")
oc1$formulas
```

## Observed confounder 2

`P` causes `D`. The sensitivity parameter is the residual `D`–`P` association.

```{r oc2}
oc2 <- placebo_lm(dat, "Y", "D", "P", covariates = "X",
                  structure = "observed_confounder_2")
oc2$formulas
```

## Post-outcome

`P` is a descendant of `Y` — reverse causation. This is the structure that fits
classical measurement-error settings, where the placebo is a noisy downstream
reading of the outcome.

```{r po2}
pox <- placebo_lm(dat, "Y", "D", "P", covariates = "X",
                  structure = "post_outcome")
pox$formulas
```

## Mediators are not supported

If `P` sits on a causal path from `D` to `Y`, this package will refuse to fit —
whether you name the structure or describe it with edges:

```{r mediator, error = TRUE}
placebo_lm(dat, "Y", "D", "P", structure = "mediator")
```

```{r mediator-edges, error = TRUE}
placebo_lm(dat, "Y", "D", "P", edges = c("D->P", "P->Y"))
```

This is deliberate. The paper is explicit that in the mediator case "the
placebo approach becomes very complicated, and we do not recommend our
approach." Returning a number here would give the appearance of an answer the
method does not support.

## Triangulating across structures

When a placebo can be read in more than one way, fitting both and comparing is
informative — the paper does exactly this in its NSW analysis using two
different placebo years. The scale factors differ, so `m` is not comparable
across them, but `k` is:

```{r compare}
fits <- list(`placebo outcome`       = po,
             `observed confounder 1` = oc1)

data.frame(
  structure = names(fits),
  SF        = sapply(fits, `[[`, "SF"),
  at_k_1    = sapply(fits, plm_estimate, k = 1),
  tipping_k = sapply(fits, function(f) plm_solve(f, 0)$k),
  row.names = NULL
)
```

Where ranges of `k` that you can defend for each reading overlap, the
intersection of the implied effect ranges is a tighter — and still
defensible — conclusion than either alone.

## Reference

Rohde, A. and Hazlett, C. Causal progress with imperfect placebo treatments and
outcomes.
